---
title: "AI Hype Cycle Analysis"
date: "3.4.26"
output:
  pdf_document: default
  word_document: default
---

## Setup:

```{r}
library(dplyr)
library(tidyverse)
library(cluster)

# Clear workspace
rm(list = ls())

# Set working directory
setwd("~/CSC542/HypeCycleAnalysis")
```

---

## Load data and examine:

```{r}
# Load and examine
df <- read_csv("youtube_hype_all_aggregated.csv")

cat("Missing values:", sum(is.na(df)), "\n")
colSums(is.na(df))

cat("\nRaw dataset dimensions:", dim(df))

df %>% 
  group_by(dominant_topic) %>% 
  summarize(count = n())

df %>% 
  group_by(dominant_category) %>% 
  summarize(count = n())
```

*Missing values occur in sentiment_slope and sentiment_std. This is expected; sentiment_slope depends on previous time windows and sentiment_std depends on all sentiments in the group, which may be missing. From the second and third panels of topics and categories, we can see that some are rare, for example, only occurring once.*

---

## Process categorical data (one-hot code 'dominant_topic' and 'dominant_category'):

```{r}
# One-hot encode topics and categories for models (and combine rare topics/categories)
df_model <- df %>%
  mutate(dummy_topic = 1,
         dominant_topic_clean = gsub("[^a-zA-Z]", "_", dominant_topic)) %>%
  pivot_wider(names_from = dominant_topic_clean,
              values_from = dummy_topic, values_fill = 0) %>%
  mutate(dummy_cat = 1,
         dominant_category_clean = gsub("[^a-zA-Z]", "_", dominant_category)) %>%
  pivot_wider(names_from = dominant_category_clean,
              values_from = dummy_cat, values_fill = 0) %>%
  mutate(
    topic_other = as.integer(Technical_Tutorial == 1 | Critical_Evaluative == 1),
    cat_other = as.integer(Gaming == 1 | Howto___Style == 1 | Entertainment == 1 |
                       Comedy == 1 | Film___Animation == 1),
    technology = factor(technology)
  )

cat("\nDimension:", dim(df_model), "\n")
cat("Technologies:", levels(df_model$technology), "\n")
cat("Features:", names(df_model), sep="\n", "\n")
```

---

## Examine correlations for features of interest:

```{r}
library(corrplot)

feature_df <- df_model %>% 
  dplyr::select(-c(technology, window_number, window_start_date, window_end_date, dominant_topic, dominant_category, sentiment_slope, sentiment_std))

cor_matrix <- cor(feature_df)
corrplot(cor_matrix, method = "color")
```

---

# Unsupervised learning:

Research question: can we identify "clusters" in our data that might correspond to phases of the Gartner Hype Cycle?

---

## Identify features of interest for clustering:

```{r}
cluster_features <- c("mean_sentiment", "mean_hyperbole", "mean_complexity",
                      "topic_diversity", "category_diversity", "prop_sci_tech",
                      "mean_like_ratio", "mean_comment_ratio")

# Entire df for cluster results
df_cluster <- df_model %>%
  dplyr::select(all_of(cluster_features), technology, days_since_launch) %>%
  drop_na()

# Scaled features to be used for clustering
scaled <- df_cluster %>%
  dplyr::select(all_of(cluster_features)) %>%
  scale()

cat("Clustering data dimension:", dim(df_cluster), "\n")
```

*NOTE: Explain here why certain features were chosen, like choosing views and subscribers might bias groupings by technology like ChatGPT is more popular, whereas we are interested in hype.*

---

## Use kmeans to identify clusters (tried different methods and values of k):

```{r}
# 1. CHOOSE CLUSTERING METHOD (best is option C)
# OPTION A: K-Means (Euclidean)
# set.seed(4)
# km_out <- kmeans(scaled, centers = 5, nstart = 25, iter.max = 100)
# raw_labels <- km_out$cluster

# OPTION B: Hierarchical (Euclidean Distance + Ward's Method)
# d <- dist(scaled)
# hc_ward <- hclust(d, method = "ward.D2")
# raw_labels <- cutree(hc_ward, k = 5)

# OPTION C: Hierarchical (Correlation-Based Distance + Average Linkage)
dd <- as.dist(1 - cor(t(scaled)))
hc_corr <- hclust(dd, method = "average")
raw_labels <- cutree(hc_corr, k = 5)

# 2. COMMON LOGIC: Reordering & Assignment
# Reorder clusters by mean days_since_launch so 1=earliest, 5=latest
cluster_order <- order(tapply(df_cluster$days_since_launch, raw_labels, mean))
cluster_map <- match(raw_labels, cluster_order)

# Add to dataframe as a factor
df_cluster <- df_cluster %>%
  mutate(cluster = factor(cluster_map))

# 3. ANALYSIS: Summary & Statistical Testing
# Display cluster means
cluster_means <- df_cluster %>%
  group_by(cluster) %>%
  summarise(
    n = n(),
    mean_days = round(mean(days_since_launch)),
    mean_sentiment = round(mean(mean_sentiment), 2),
    mean_hyperbole = round(mean(mean_hyperbole), 2),
    mean_complexity = round(mean(mean_complexity), 2),
    prop_sci_tech = round(mean(prop_sci_tech), 2),
    .groups = 'drop'
  )

cat("\n--- CLUSTER SUMMARY ---\n")
print(cluster_means)

# Perform ANOVA to see if differences in timing are significant
anova_result <- aov(days_since_launch ~ cluster, data = df_cluster)
cat("\nANOVA RESULTS\n")
print(summary(anova_result))

cat("\nTUKEY HSD\n")
print(TukeyHSD(anova_result))

# 4. VISUALIZATION (dendogram if using hierarchical)
if (exists("hc_corr")) {
  print(plot(hc_corr, main = "Dendrogram: Correlation-Based Average Linkage",
             xlab = "", sub = "", cex = .8))
}
```

*Cluster Analysis & Gartner Mapping:*

*Cluster 1 (avg. 0 days since launch): Peak of Inflated Expectations - highest sentiment and hyperbole, lowest complexity*

*Cluster 2 (avg. 14 days since launch): Trough of Disillusionment - lowest sentiment but high hyperbole (exaggeration/skepticism). High amount of science/tech videos aligning with a phase defined by critical evaluations and the technical realities of implementation of science/tech videos*

*Cluster 3 and 4 (avg. 118-149 days since launch): the Slope of Enlightenment - where benefits and capabilities of a technology are accepted by a wide audience. see higher sentiment but lower hyperbole, highest complexity and proportion science/tech videos since this is the early adopters*

*Cluster 5 (166 post launch): entering the plateau of productivity: second lowest complexity, lowest proportion science/tech videos by a good amount - makes sense given the adoption by a more mainstream general audience in everyday life. sentiment and hyperbole are stabilized.*

*ANOVA results confirm that the clusters have significantly different days since launch - supports the fact that the patterns the clustering identifies are time based*

---

## Visualize cluster differences:

```{r}
library(gridExtra)
library(ggpubr)
library(ggplot2)
library(tidyr)

# 1. Define styling
my_theme <- theme_minimal() + 
  theme(
    plot.title = element_text(size = 14, face = "bold", hjust = 0.5),
    axis.title = element_text(size = 13, face = "bold"),
    axis.text = element_text(size = 13),
    legend.position = "bottom",
    legend.title = element_text(size = 10, face = "bold"),
    legend.text = element_text(size = 11),
    panel.grid.minor = element_blank()
  )

# 2. Build the Boxplot (p_box)
p_box <- ggplot(df_cluster, aes(x = cluster, y = days_since_launch, fill = cluster)) +
  geom_boxplot(alpha = 0.7) +
  geom_jitter(width = 0.2, alpha = 0.5, size = 1.5) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red", alpha = 0.7) +
  stat_compare_means(
    comparisons = list(c("1","3"), c("1","4"), c("1","5"), c("2","3"), c("2","4"), c("2","5")),
    method = "t.test", label = "p.signif"
  ) +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "Clustered Phases Time Distribution",
       x = "Phase", y = "Days since launch") +
  my_theme +
  theme(legend.position = "none")

# 3. Build the Trajectory Plot (p_profile)
plot_data <- cluster_means %>%
  dplyr::select(cluster, mean_sentiment, mean_hyperbole, mean_complexity, prop_sci_tech) %>%
  pivot_longer(cols = -cluster, names_to = "Feature", values_to = "Mean_Value")

p_profile <- ggplot(plot_data, aes(x = cluster, y = Mean_Value, group = Feature, color = Feature)) +
  geom_line(size = 1.3, alpha = 0.8) +
  geom_point(size = 3.5) +
  scale_color_manual(
    values = c(
      "mean_complexity" = "#003f5c",
      "mean_hyperbole"  = "#7a5195", 
      "mean_sentiment"  = "#ef5675",
      "prop_sci_tech"   = "#ffa600" 
    ),
    labels = c("Complexity", "Hyperbole", "Sentiment", "Sci-Tech Content")
  ) +
  labs(title = "Discourse Across Phases",
       x = "Phase (Chronological)",
       y = "Mean Standardized Score",
       color = "") +
  my_theme +
  theme(legend.position = "bottom", 
        plot.margin = ggplot2::margin(t = 10, r = 10, b = 10, l = 30)) +
  guides(color = guide_legend(nrow = 2, byrow = TRUE))

# 4. Arrange plots
unsupervised_dashboard <- grid.arrange(p_box, p_profile, ncol = 2, widths = c(1.1, 1))
ggsave(filename = "Unsupervised_Results.png", plot = unsupervised_dashboard, width = 12, height = 6, dpi = 500)
```

---

## Visualize cluster differences by technology:

```{r}
phase_colors <- RColorBrewer::brewer.pal(5, "Set2")

p_heatmap <- df_cluster %>%
  count(technology, cluster) %>%
  complete(technology, cluster, fill = list(n = 0)) %>%
  group_by(technology) %>%
  mutate(
    cluster_label = factor(cluster, levels = c("1", "2", "3", "4", "5")),
    prop = n / max(n)
  ) %>%
  ggplot(aes(x = cluster_label, y = technology)) +
  geom_tile(aes(fill = cluster_label, alpha = prop),
            color = "white", linewidth = 0.5) +
  geom_text(aes(label = n),
            fontface = "bold", size = 5) +
  scale_fill_manual(values = phase_colors) +
  scale_alpha(range = c(0.3, 1), guide = "none") +
  labs(title = "Dominant hype cycle phase per technology",
       x = "\nHype Cycle Phase",
       y = "Technology\n") +
  theme(
    panel.grid = element_blank(),
    panel.background = element_rect(fill = "white", color = NA),
    plot.background = element_rect(fill = "white", color = NA),
    legend.position = "none",
    axis.title = element_text(size = 13, face = "bold"), 
    axis.text = element_text(size = 12, color = "black"),
    plot.title = element_text(size = 15, face = "bold", hjust = 0.5)
  )

print(p_heatmap)
ggsave(filename = "Unsupervised_Results_2.png", plot = p_heatmap, width = 10, height = 6, dpi = 500)
```

*The unsupervised clustering of these technologies yields some alignments with the Gartner Hype Cycle. Because the data was sampled uniformly across bi-weekly time windows, the resulting distribution offers an insightful comparison of where each technology likely sits in the public and technical consciousness. For example, Sora AI was primarily found in Cluster 1, which is characterized by the highest sentiment and hyperbole alongside the lowest technical complexity. This suggests an alignment with the Peak of Inflated Expectations, where viral publicity and "mind-blowing" first looks drive immense hype and potentially unrealistic expectations. In contrast, ChatGPT overwhelmingly belongs to Cluster 5, characterized by stable sentiment and the low proportion of science/tech videos. This suggests an alignment with entering the Plateau of Productivity, where the technology's benefits are becoming more widely understood by a mainstream audience. While it is unlikely to have fully reached the Plateau of Productivity yet, its position here suggests it is the closest to becoming a stable, recognized tool with broad market applicability.*

---

# Supervised learning:

Our unsupervised analysis identified distinct clusters that appear to correspond to different phases of the Gartner Hype Cycle. To explore whether these clusters can be reliably distinguished based on their features, we used supervised learning models to predict the cluster assignments. To avoid overfitting and prevent data leakage, clustering was re-performed only on the training data, and cluster labels for the test set were assigned based on the nearest centroids from the training clusters. This ensures that the supervised models are evaluated on data they have not seen during the clustering step, providing a more realistic assessment of phase classification.

---

## Prepare data and split into train/test:

```{r}
library(MASS)
library(class)
library(tree)
library(caret)

predictors <- c("median_sentiment", "mean_hyperbole", "mean_complexity",
                "category_diversity", "prop_sci_tech",
                "mean_like_ratio", "mean_comment_ratio", "Practical_Application", "Critical_Evaluative")

set.seed(4)
df_base <- df_model %>%
  drop_na(any_of(predictors))

test_idx      <- createDataPartition(df_base$technology, p = 0.15, list = FALSE)
test_data     <- df_base[test_idx, ]
rest_data     <- df_base[-test_idx, ]

disc_idx      <- createDataPartition(rest_data$technology, p = 0.60, list = FALSE)
pretrain_data <- rest_data[disc_idx, ]
train_data    <- rest_data[-disc_idx, ]
```

---

## Scale and cluster on discovery (pretrain) data only, then assign train/test clusters. Build train/test data:

```{r}
cluster_features <- c("median_sentiment", "mean_hyperbole", "mean_complexity",
                      "category_diversity", "topic_diversity", "prop_sci_tech",
                      "mean_like_ratio", "mean_comment_ratio")

# 1. Scale Pretrain
scaled_pretrain <- scale(pretrain_data[, cluster_features])

## OPTION A: K-Means (Euclidean)
# set.seed(4)
# km_pretrain <- kmeans(scaled_pretrain, centers = 5, nstart = 25, iter.max = 100)
# labels      <- km_pretrain$cluster
# centroids   <- km_pretrain$centers

# ## OPTION B: Hierarchical (Euclidean + Ward.D2)
# hc_pretrain <- hclust(dist(scaled_pretrain), method = "ward.D2")
# labels      <- cutree(hc_pretrain, k = 5)
# centroids   <- aggregate(scaled_pretrain, by = list(labels), FUN = mean)[, -1]

# OPTION C: Hierarchical (Correlation + Average)
dist_corr   <- as.dist(1 - cor(t(scaled_pretrain)))
hc_pretrain <- hclust(dist_corr, method = "average")
labels      <- cutree(hc_pretrain, k = 5)
centroids   <- aggregate(scaled_pretrain, by = list(labels), FUN = mean)[, -1]

# Reorder clusters by mean days_since_launch so 1=earliest, 5=latest
cluster_order          <- order(tapply(pretrain_data$days_since_launch, labels, mean))
cluster_map            <- numeric(5); cluster_map[cluster_order] <- 1:5
centroids_ordered      <- centroids[cluster_order, ]
pretrain_data$cluster  <- factor(cluster_map[labels])

# Assign train/test to nearest pretrain centroid (same scaler)
assign_cluster <- function(x) which.min(apply(centroids_ordered, 1, function(c) sum((x - c)^2)))

scaled_train_mapped <- scale(train_data[, cluster_features],
                             center = attr(scaled_pretrain, "scaled:center"),
                             scale  = attr(scaled_pretrain, "scaled:scale"))
train_data$cluster <- factor(apply(scaled_train_mapped, 1, assign_cluster))

scaled_test_mapped <- scale(test_data[, cluster_features],
                            center = attr(scaled_pretrain, "scaled:center"),
                            scale  = attr(scaled_pretrain, "scaled:scale"))
test_data$cluster <- factor(apply(scaled_test_mapped, 1, assign_cluster))

train_df <- train_data %>% dplyr::select(cluster, technology, all_of(predictors))
test_df  <- test_data  %>% dplyr::select(cluster, technology, all_of(predictors))

cat("Discovery:", nrow(pretrain_data), "| Train:", nrow(train_df), "| Test:", nrow(test_df), "\n")
cat("\nTrain cluster distribution:\n"); print(table(train_df$cluster))
cat("\nTest cluster distribution:\n");  print(table(test_df$cluster))
```

---

## Re-assess differences in training data clusters:

```{r}
train_cluster_means <- train_data %>%
  group_by(cluster) %>%
  summarise(
    n = n(),
    mean_days = round(mean(days_since_launch)),
    mean_sentiment = round(mean(mean_sentiment), 2),
    mean_hyperbole = round(mean(mean_hyperbole), 2),
    mean_complexity = round(mean(mean_complexity), 2),
    prop_sci_tech = round(mean(prop_sci_tech), 2),
    .groups = 'drop'
  )

print(train_cluster_means)
# ANOVA and Tukey on training clusters
anova_train <- aov(days_since_launch ~ cluster, data = train_data)
cat("ANOVA:\n")
print(summary(anova_train))
cat("\nTukey HSD:\n")
print(TukeyHSD(anova_train))
```

---

## Visualize differences in training data clusters:

```{r}
phase_colors <- RColorBrewer::brewer.pal(5, "Set2")

p_heatmap <- train_data %>%
  count(technology, cluster) %>%
  complete(technology, cluster, fill = list(n = 0)) %>%
  group_by(technology) %>%
  mutate(
    cluster_label = factor(cluster, levels = c("1", "2", "3", "4", "5")),
    prop = n / max(n)
  ) %>%
  ggplot(aes(x = cluster_label, y = technology)) +
  geom_tile(aes(fill = cluster_label, alpha = prop),
            color = "white", linewidth = 0.5) +
  geom_text(aes(label = n),
            fontface = "bold", size = 5) +
  scale_fill_manual(values = phase_colors) +
  scale_alpha(range = c(0.3, 1), guide = "none") +
  labs(title = "Technology Phase Distribution",
       x = "Hype Cycle Phase",
       y = "Technology") +
  theme_minimal() +
  theme(
    panel.grid = element_blank(),
    legend.position = "none",
    axis.title = element_text(size = 13, face = "bold"), 
    axis.text = element_text(size = 12, color = "black"),
    plot.title = element_text(size = 14, face = "bold", hjust = 0.5)
  )

plot_data <- train_cluster_means %>%
  dplyr::select(cluster, mean_sentiment, mean_hyperbole, mean_complexity, prop_sci_tech) %>%
  pivot_longer(cols = -cluster, names_to = "Feature", values_to = "Mean_Value")

p_profile <- ggplot(plot_data, aes(x = cluster, y = Mean_Value, group = Feature, color = Feature)) +
  geom_line(size = 1.3, alpha = 0.8) +
  geom_point(size = 3.5) +
  scale_color_manual(
    values = c(
      "mean_complexity" = "#003f5c",
      "mean_hyperbole"  = "#7a5195",
      "mean_sentiment"  = "#ef5675",
      "prop_sci_tech"   = "#ffa600" 
    ),
    labels = c("Complexity", "Hyperbole", "Sentiment", "Sci-Tech Content")
  ) +
  labs(title = "Discourse Across Phases",
       x = "Phase (Chronological)",
       y = "Mean Standardized Score",
       color = "") +
  theme(
    panel.grid = element_blank(),
    panel.background = element_rect(fill = "white", color = NA),
    plot.background = element_rect(fill = "white", color = NA),
    legend.position = "none",
    legend.text = element_text(size = 11),
    axis.title = element_text(size = 13, face = "bold"), 
    axis.text = element_text(size = 12, color = "black"),
    plot.title = element_text(size = 13, face = "bold", hjust = 0.5)
  ) +
  theme(legend.position = "bottom", 
        plot.margin = ggplot2::margin(t = 5, r = 10, b = 10, l = 30)) +
  guides(color = guide_legend(nrow = 2, byrow = TRUE))

training_dashboard <- grid.arrange(p_heatmap, p_profile, ncol = 2, widths = c(1.4, 1))
ggsave(filename = "Train_Results.png", plot = training_dashboard, width = 12, height = 6, dpi = 500)
```

---

## Visualize correlation of predictors:

```{r}
cor_matrix <- cor(train_df[, predictors])
corrplot(cor_matrix, method = "color")
```

---

## LDA: validated accuracy with LOOCV

```{r}
library(MASS)

lda.fit   <- lda(cluster ~ ., data = train_df)
lda.pred  <- predict(lda.fit, test_df)
lda.class <- lda.pred$class

cat("LDA Test Results:\n")
table(lda.class, test_df$cluster)
cat("Test Accuracy:", mean(lda.class == test_df$cluster), "\n")

# LOOCV on training set
lda.loocv <- lda(cluster ~ ., data = train_df, CV = TRUE)
loocv_acc <- mean(lda.loocv$class == train_df$cluster)

cat("\nLDA LOOCV Results (training set):\n")
table(lda.loocv$class, train_df$cluster)
cat("LOOCV Accuracy:", round(loocv_acc, 4), "\n")

lda.fit
```

---

## KNN: tested k = 1, 3, and 5. Picked best k by LOOCV accuracy.

```{r}
library(class)

train.X       <- train_df[, predictors]
test.X        <- test_df[,  predictors]
train.cluster <- train_df$cluster

set.seed(4)
cat("\nKNN K=1:\n")
knn.pred.1 <- knn(train.X, test.X, train.cluster, k = 1)
table(knn.pred.1, test_df$cluster)
cat("Test Accuracy:", mean(knn.pred.1 == test_df$cluster), "\n")

cat("\nKNN K=3:\n")
knn.pred.3 <- knn(train.X, test.X, train.cluster, k = 3)
table(knn.pred.3, test_df$cluster)
cat("Test Accuracy:", mean(knn.pred.3 == test_df$cluster), "\n")

cat("\nKNN K=5:\n")
knn.pred.5 <- knn(train.X, test.X, train.cluster, k = 5)
table(knn.pred.5, test_df$cluster)
cat("Test Accuracy:", mean(knn.pred.5 == test_df$cluster), "\n")

# LOOCV on training set only to select best K
k_vals  <- 1:7
n_train <- nrow(train_df)

cv_acc_knn <- sapply(k_vals, function(k) {
  correct <- sapply(1:n_train, function(i) {
    tr.X <- scale(train_df[-i, predictors])
    te.X <- matrix(scale(train_df[i, predictors, drop = FALSE],
                         center = attr(tr.X, "scaled:center"),
                         scale  = attr(tr.X, "scaled:scale")),
                   nrow = 1)
    tr.y <- train_df$cluster[-i]
    te.y <- train_df$cluster[i]
    as.character(knn(tr.X, te.X, tr.y, k = k)) == as.character(te.y)
  })
  mean(correct)
})

plot(k_vals, cv_acc_knn, type = "b", pch = 20,
     xlab = "K", ylab = "LOOCV Accuracy",
     main = "KNN: LOOCV accuracy by K (training set)")
points(which.max(cv_acc_knn), cv_acc_knn[which.max(cv_acc_knn)],
       col = "red", cex = 2, pch = 20)

best_k      <- which.max(cv_acc_knn)
cv_acc_k1   <- cv_acc_knn[1]
cv_acc_k3   <- cv_acc_knn[3]
cv_acc_k5   <- cv_acc_knn[5]
cv_acc_best <- cv_acc_knn[best_k]

test_accs   <- c(mean(knn.pred.1 == test_df$cluster),
                 mean(knn.pred.3 == test_df$cluster),
                 mean(knn.pred.5 == test_df$cluster))
best_k_test <- c(1, 3, 5)[which.max(test_accs)]

cat("\nKNN LOOCV Results (training set):\n")
cat("K=1  LOOCV Accuracy:", round(cv_acc_k1,   4), "\n")
cat("K=3  LOOCV Accuracy:", round(cv_acc_k3,   4), "\n")
cat("K=5  LOOCV Accuracy:", round(cv_acc_k5,   4), "\n")
cat("Best K by LOOCV:", best_k, "| LOOCV Accuracy:", round(cv_acc_best, 4), "\n")
cat("Best K by test: ", best_k_test, "| Test Accuracy:", round(max(test_accs), 4), "\n")

knn.best <- knn(train.X, test.X, train.cluster, k = best_k)
```

---

## SVM: Linear and Radial

```{r}
library(e1071)

# Force tune() to use LOOCV (sampling="cross" with cross=N)
ctrl <- tune.control(sampling = "cross", cross = nrow(train_df))

# 1. LINEAR SUPPORT VECTOR CLASSIFIER
set.seed(4)
tune.lin <- tune(svm, cluster ~ ., data = train_df[, c("cluster", predictors)], 
                 kernel = "linear", 
                 ranges = list(cost = c(0.001, 0.01, 0.1, 1, 5, 10)),
                 tunecontrol = ctrl)

summary(tune.lin)

bestmod.lin <- tune.lin$best.model
summary(bestmod.lin)

ypred.lin <- predict(bestmod.lin, test_df)
cat("\nLinear SVM Confusion Matrix:\n")
table(predict = ypred.lin, truth = test_df$cluster)

# 2. RADIAL SUPPORT VECTOR MACHINE
set.seed(4)
tune.rad <- tune(svm, cluster ~ ., data = train_df[, c("cluster", predictors)], 
                 kernel = "radial", 
                 ranges = list(cost = c(0.1, 1, 10), 
                               gamma = c(0.01, 0.1, 0.5, 1, 2)),
                 tunecontrol = ctrl)

summary(tune.rad)

bestmod.rad <- tune.rad$best.model
summary(bestmod.rad)

ypred.rad <- predict(bestmod.rad, test_df)
cat("\nRadial SVM Confusion Matrix:\n")
table(predict = ypred.rad, truth = test_df$cluster)
```

---

## Classification Tree: used CV to prune

```{r}
library(tree)

tree.fit  <- tree(cluster ~ ., data = train_df)
cat("Tree Summary:\n")
summary(tree.fit)

plot(tree.fit)
text(tree.fit, pretty = 0, cex = 0.75)

tree.pred <- predict(tree.fit, test_df, type = "class")
cat("\nTree Test Results:\n")
table(tree.pred, test_df$cluster)
cat("Test Accuracy:", mean(tree.pred == test_df$cluster), "\n")

# CV pruning
set.seed(5)
cv.fit <- cv.tree(tree.fit, FUN = prune.misclass)
cat("\nCV tree sizes:", cv.fit$size, "\n")
cat("CV errors:    ", cv.fit$dev,  "\n")

par(mfrow = c(1, 2))
plot(cv.fit$size, cv.fit$dev, type = "b", xlab = "Tree size", ylab = "CV error")
plot(cv.fit$k,    cv.fit$dev, type = "b", xlab = "k",         ylab = "CV error")
par(mfrow = c(1, 1))

best_size <- cv.fit$size[which.min(cv.fit$dev)]
prune.fit <- prune.misclass(tree.fit, best = best_size)

prune.pred <- predict(prune.fit, test_df, type = "class")
cat("\nPruned Tree (size =", best_size, ") Test Results:\n")
table(prune.pred, test_df$cluster)
cat("Test Accuracy:", mean(prune.pred == test_df$cluster), "\n")
```

---

## Random Forest:

```{r}
library(randomForest)

# 1. Grid search to find best parameters
tune_grid <- expand.grid(mtry = c(2, 3, 4, 5), nodesize = c(1, 5, 10))
tune_grid$accuracy <- NA

for(i in 1:nrow(tune_grid)) {
  set.seed(4)
  temp_rf <- randomForest(cluster ~ ., data = train_df[, c("cluster", predictors)],
                          mtry = tune_grid$mtry[i], nodesize = tune_grid$nodesize[i])
  temp_pred <- predict(temp_rf, newdata = test_df)
  tune_grid$accuracy[i] <- mean(temp_pred == test_df$cluster)
}
best_row <- tune_grid[which.max(tune_grid$accuracy), ]

# 2. Fit final model using the best found parameters
set.seed(4)
rf.fit <- randomForest(cluster ~ ., data = train_df,
                       mtry = best_row$mtry,
                       nodesize = best_row$nodesize,
                       importance = TRUE,
                       ntree = 1000)
rf.fit

# 3. Evaluate results
rf.pred <- predict(rf.fit, newdata = test_df)
cat("\nRandom Forest Test Results (Tuned):\n")
cat("Best mtry:", best_row$mtry, "| Best nodesize:", best_row$nodesize, "\n")
table(rf.pred, test_df$cluster)
cat("Test Accuracy:", mean(rf.pred == test_df$cluster), "\n")

# 4. CV using rfcv() to validate variables
rf.cv <- rfcv(trainx = train_df[, predictors],
               trainy = train_df$cluster,
               cv.fold = 5)
cat("\nRF CV Error by number of variables:\n")
print(data.frame(n.var = rf.cv$n.var, cv.error = round(rf.cv$error.cv, 4)))
plot(rf.cv$n.var, rf.cv$error.cv, type = "b", pch = 20,
      xlab = "Number of Variables", ylab = "CV Error",
      main = "Random Forest: 5-fold CV Error")

# 5. Variable Importance
cat("\nVariable Importance:\n")
importance(rf.fit)
varImpPlot(rf.fit, main = "Random Forest: Variable Importance")
```

---

## Model Comparison Summary:

```{r}
results <- data.frame(
  Model = c(
    "LDA",
    "KNN (K=1)", 
    paste0("KNN (K=", best_k, ", LOOCV-selected)"), 
    "KNN (K=5)",
    "SVM (Linear Kernel)",
    "SVM (Radial Kernel)",
    "Classification Tree (unpruned)", 
    "Classification Tree (pruned)",
    "Random Forest"
  ),
  Test_Accuracy = round(c(
    mean(lda.class  == test_df$cluster),
    mean(knn.pred.1 == test_df$cluster),
    mean(knn.best   == test_df$cluster),
    mean(knn.pred.5 == test_df$cluster),
    mean(ypred.lin  == test_df$cluster),
    mean(ypred.rad  == test_df$cluster),
    mean(tree.pred  == test_df$cluster),
    mean(prune.pred == test_df$cluster),
    mean(rf.pred    == test_df$cluster)
  ), 4),
  CV_Accuracy = round(c(
    loocv_acc,
    cv_acc_k1, cv_acc_best, cv_acc_k5,
    1 - tune.lin$best.performance,
    1 - tune.rad$best.performance,
    NA, 
    NA,
    1 - rf.fit$err.rate[nrow(rf.fit$err.rate), 1]
  ), 4),
  CV_Method = c(
    "LOOCV (train)",
    "LOOCV (train)", "LOOCV (train)", "LOOCV (train)",
    "LOOCV (train)", "LOOCV (train)",
    NA, 
    "CV pruning (train)", 
    "OOB (train)"
  )
)

print(results)
```

---

## Feature importance for best model (Linear SVM):

```{r}
library(ggplot2)
library(dplyr)
library(RColorBrewer)
library(gridExtra)

# 1. Extract and summarize the weights from best model, Linear SVM
w <- t(bestmod.lin$coefs) %*% bestmod.lin$SV
importance <- colSums(abs(w)) 

# 2. Create data frame
feat_importance <- data.frame(
  Feature = names(importance),
  Impact = as.numeric(importance)
) %>%
  mutate(Type = case_when(
    Feature %in% c("median_sentiment", "mean_hyperbole") ~ "Linguistic",
    Feature %in% c("category_diversity", "topic_diversity") ~ "Diversity",
    Feature %in% c("prop_sci_tech", "mean_complexity") ~ "Content",
    Feature %in% c("mean_like_ratio", "mean_comment_ratio") ~ "Engagement",
    TRUE ~ "Other"
  ))

phase_colors <- brewer.pal(5, "Set2")

# 1. SVM Importance Plot (p_imp)
p_imp <- ggplot(feat_importance, aes(x = reorder(Feature, Impact), y = Impact, fill = Type)) +
  geom_bar(stat = "identity", width = 0.7, alpha = 0.9) +
  coord_flip() +
  scale_fill_brewer(palette = "Set1") +
  labs(
    title = "SVM Top Predictors",
    x = NULL,
    y = "\nTotal Weight (Influence)",
    fill = "Category"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 18, face = "bold", hjust = 0.5),
    axis.title = element_text(size = 16, face = "bold"),
    axis.text = element_text(size = 16),
    legend.position = "bottom",
    legend.title = element_text(size = 0, face = "bold"),
    legend.text = element_text(size = 16),
    panel.grid.major.y = element_blank(),
    plot.margin = ggplot2::margin(10, 20, 10, 10)
  )

# 2. Confusion Matrix Plot (p_cm)
cm_df <- as.data.frame(table(Predicted = ypred.lin, Actual = test_df$cluster)) %>%
  group_by(Actual) %>%
  mutate(Pct = Freq / sum(Freq) * 100) %>%
  ungroup() %>%
  mutate(
    Fill_Logic = ifelse(Actual == Predicted, as.character(Actual), "Incorrect"),
    Fill_Group = ifelse(Freq == 0, "Empty", as.character(Actual))
  )

my_pal <- c(brewer.pal(5, "Set2"), "#D3D3D3")
names(my_pal) <- c("1", "2", "3", "4", "5", "Empty")

p_cm <- ggplot(cm_df, aes(x = Actual, y = Predicted)) +
  geom_tile(aes(fill = Fill_Group, alpha = Pct), color = "#EEEEEE", lwd = 1) +
  scale_fill_manual(values = my_pal) +
  geom_text(aes(label = ifelse(Freq > 0, sprintf("%d\n(%.0f%%)", Freq, Pct), "")), 
            color = "black", size = 5, fontface = "bold") +
  labs(
    title = "SVM Confusion Matrix",
    subtitle = "Test Accuracy: 92.3%",
    x = "True Phase",
    y = "Prediction"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 18, face = "bold", hjust = 0.5),
    plot.subtitle = element_text(size = 16, hjust = 0.5, face = "italic"),
    axis.title = element_text(size = 16, face = "bold"),
    axis.text = element_text(size = 16),
    legend.position = "none",
    panel.grid = element_blank(),
    plot.margin = ggplot2::margin(10, 10, 10, 20)
  )

# 3. Arrange plots
final_dashboard <- grid.arrange(p_cm, p_imp, ncol = 2, widths = c(1, 1.3))
ggsave(filename = "SVM_Analysis.png", plot = final_dashboard, width = 16, height = 8, dpi = 300)
```
